On the Gravitational stability of gravito-turbulent accretion disks

### 2016-01-28 paper/model3d.tex
revision 3 by MKL

\section{Basic equations}\label{basic}
We consider a 3D, self-gravitating,
viscous disk with heating and cooling. We use the shearing box framework to
study a small patch of the disk \citep{goldreich65}. The local frame
co-rotates with a fiducial point in the unperturbed disk at angular
frequency $\Omega$. The Cartesian co-ordinates $(x,y,z)$ correspond
to the radial, azimuthal and vertical directions in the global disk. 
The fluid equations are
\begin{align}
  &\frac{\p\rho}{\p t} + \nabla\cdot\left(\rho\bm{v}\right) = 0, \\
  & \frac{\p\bm{v}}{\p t} + \bm{v}\cdot\nabla\bm{v} =
  -\frac{1}{\rho}\nabla P - \nabla\Phi - 2\Omega\hat{\bm{z}}\times\bm{v}\notag\\ & 
  \phantom{\frac{\p\bm{v}}{\p t} + \bm{v}\cdot\nabla\bm{v} =}
  +2\Omega^2qx\hat{\bm{x}} -\Omega_z^2z\hat{\bm{z}} 
   + \frac{1}{\rho}\nabla\cdot\bm{T},\label{momentum_eq}\\
  & \frac{\p E}{\p t} + \nabla\cdot\left(E\bm{v}\right) = - P
  \nabla\cdot\bm{v} + \mathcal{H}_\mathrm{visc} - \Lambda +
  \mathcal{H}_\mathrm{ext} ,\label{energy_eq}
\end{align} 
where $\rho$ is the density field and $\bm{v} = (v_x, v_y, v_z)$ is
the velocity field. 
We assume an ideal gas so that the pressure $P$ and thermal energy
density $E$ are related by  
\begin{align}
  P = (\gamma-1)E = \mathcal{R}\rho T, 
\end{align}
where $\mathcal{R}$ is the gas constant and $T$ is the temperature. {\bf KMK cut because 
mostly varied: We
use $\gamma=1.4$ unless otherwise stated.} The gas 
gravitational potential $\Phi$ is given via the Poisson equation,  
\begin{align}
 \nabla^2\Phi = 4 \pi G \rho. 
\end{align}

@@

While steady-state viscosity values can be determined 
analytically or numerically \citep[e.g.][]{martin11,kratter08,rafikov15}, the time-dependent behavior is
not well-explored. We emphasize the choice $\mu=-1,\lambda=0$ is made
to bring out the physical process of interest --- viscous GI. 
Interestingly, though, \cite{laughlin96b} have suggested a $\nu \propto 1/\Sigma$
dependence when modeling the evolution of 2D self-gravitating disks 
as a viscous process. We note that if $\rho\nu$ is constant in time
then increasing the density leads to reduction in the viscosity. 
This is perhaps not inconsistent with numerical simulations of disk
fragmentation which show the internal flow of high-density clumps
is laminar \citep{gammie01}, although once clumps form they may effectively decouple
from the background disk state, and thus no longer be described by the same prescription.   
{\bf is this true? KMK: maybe caveat helps?}

@@

\subsubsection{$\beta$ cooling}\label{beta_cool_model}
In our $\beta$ cooling model, the energy loss per unit volume is specified  
as an explicit function of the thermodynamic variables. 
A prototypical example is 
\begin{align}\label{beta_cool}
  \Lambda(\rho, T) =
  %  \frac{P}{(\gamma-1)T}\frac{\left(T-T_\mathrm{irr}\right)}{t_c}, 
  \frac{\mathcal{R}\rho}{(\gamma-1)}\frac{\left(T-T_\mathrm{irr}\right)}{t_c}. 
\end{align}
Recall  $\tcool = \beta\Omega^{-1}$ is 
the cooling timescale with $\beta$ a constant input parameter. Here, 
\begin{align}\label{tirr_def}
  \tirr = \theta T_\mathrm{eq}
\end{align}
is a reference temperature field parameterized as a constant 
fraction $\theta$ of the equilibrium temperature
$T_\mathrm{eq}$. Physically, $\tirr$ is the floor temperature set by,
for example, stellar or background irradiation. We will thus refer to
$\theta$ as the `irradiation level'. As we discuss in Section ?, 
the limits $\theta \rightarrow 0, \theta \rightarrow 1$ do not exactly 
correspond to the limits of non-irradiated or irradiation-dominated disks
as discussed in the literature \citep{kratter11,rice11}.%stellar illum or envelope rad

For $\beta$-cooling given by Eq. \ref{beta_cool} we assume an 
equilibrium polytropic relation 
\begin{align} 
  P  =
%  \frac{c_{s0}^2\rho_0}{\Gamma}\left(\frac{\rho}{\rho_0}\right)^\Gamma,
  %  \equiv \frac{c_s^2(z)}{\Gamma}\rho,
c_{s0}^2\rho_0\left(\frac{\rho}{\rho_0}\right)^\Gamma,
\end{align}
where $\Gamma$ is the constant polytropic index and $\rho_0 =
\rho(z=0)$ is the equilibrium midplane density. 
The vertical structure is first obtained from
Eq. \ref{vert_eq1}---\ref{vert_eq2} then inserted into
Eq. \ref{thermal_eq} to infer the required viscosity profile for
thermal equilibrium. If  $\mathcal{H}_\mathrm{ext}=0$, 
\begin{align}\label{alpha_beta_relation}
\alpha(z) = \frac{1}{(\gamma-1)\beta
   q^2}\frac{c_s^2(z)}{c_{s0}^2}\left(1 - \theta\right). 
%  \alpha(z) = \frac{1}{\Gamma(\gamma-1)\beta
%    q^2}\left(\frac{\rho}{\rho_0}\right)^{\Gamma-1} \left(1 - \theta\right). 
\end{align}

@@

Note that, although we assume standard beta cooling in formulating the
linear problem later, any explicit 
cooling function, say $\Lambda_1(\rho,T)$, can be mapped to
that obtained from Eq. \ref{beta_cool} by equating their linearized
forms, i.e. $\delta \Lambda \equiv \delta \Lambda_1$. This then 
defines the $\beta$ and $\theta$ parameters to be used in the 
framework we develop. We do this in \S\ref{2dppd} where we adopt a
more realistic beta cooling function for PPDs. In this case, $\theta$
may or may not represent a physical irradiation.

\subsubsection{Radiative cooling}\label{rad_cool}
A more realistic treatment of cooling considers energy transfer by radiative diffusion. Then
\begin{align}
  \Lambda &= \nabla\cdot\bm{F},\label{rad_cool1}\\
  \bm{F}   &= -\frac{16\sigma T^3}{3\kappa_d\rho}\nabla T, \label{rad_cool2}
\end{align}
where $\sigma$ is the Stefan-Boltzmann constant and 
$\kappa_d$ is the (dust) opacity. We adopt
\begin{align}\label{opacity_law}
  \kappa_d = \kappa_{d0}T^b,
\end{align}
and take the constant index $b=2$ as appropriate for the cold outer regions
of a PPD with ISM-like dust grains \citep{bell94}, but retain the general notation $b$
to keep track of the opacity.    
%We use radiative cooling when considering 3D PPDs in \S.

@@

While radiative cooling is arguably more realistic than
beta cooling, it generally implies a vertically
non-isothermal equilibrium disk, and increases the order of the
linearized equations. It formally applies to optically-thick 
disks, but it is possible to modify the flux function to account
for optically-thin disks. However, this complication is beyond the scope of
this work. {\bf ref for flux lim diff? \citep{levermore81}?}

>>>>

\section{Basic equations}\label{basic}
We consider a 3D, self-gravitating,
viscous disk with heating and cooling. We use the shearing box framework to
study a small patch of the disk \citep{goldreich65}. The local frame
co-rotates with a fiducial point in the unperturbed disk at angular
frequency $\Omega$. The Cartesian co-ordinates $(x,y,z)$ correspond
to the radial, azimuthal and vertical directions in the global disk. 
The fluid equations are
\begin{align}
  &\frac{\p\rho}{\p t} + \nabla\cdot\left(\rho\bm{v}\right) = 0, \\
  & \frac{\p\bm{v}}{\p t} + \bm{v}\cdot\nabla\bm{v} =
  -\frac{1}{\rho}\nabla P - \nabla\Phi - 2\Omega\hat{\bm{z}}\times\bm{v}\notag\\ & 
  \phantom{\frac{\p\bm{v}}{\p t} + \bm{v}\cdot\nabla\bm{v} =}
  +2\Omega^2qx\hat{\bm{x}} -\Omega_z^2z\hat{\bm{z}} 
   + \frac{1}{\rho}\nabla\cdot\bm{T},\label{momentum_eq}\\
  & \frac{\p E}{\p t} + \nabla\cdot\left(E\bm{v}\right) = - P
  \nabla\cdot\bm{v} + \mathcal{H}_\mathrm{visc} - \Lambda +
  \mathcal{H}_\mathrm{ext} ,\label{energy_eq}
\end{align} 
where $\rho$ is the density field and $\bm{v} = (v_x, v_y, v_z)$ is
the velocity field. 
We assume an ideal gas so that the pressure $P$ and thermal energy
density $E$ are related by  
\begin{align}
  P = (\gamma-1)E = \mathcal{R}\rho T, 
\end{align}
where $\mathcal{R}$ is the gas constant and $T$ is the temperature. The gas 
gravitational potential $\Phi$ is given via the Poisson equation,  
\begin{align}
 \nabla^2\Phi = 4 \pi G \rho. 
\end{align}

@@

While steady-state viscosity values can be determined 
analytically or numerically \citep[e.g.][]{martin11,kratter08,rafikov15}, the time-dependent behavior is
not well-explored. We emphasize the choice $\mu=-1,\lambda=0$ is made
to bring out the physical process of interest --- viscous GI. 
Interestingly, though, \cite{laughlin96b} have suggested a $\nu \propto 1/\Sigma$
dependence when modeling the evolution of 2D self-gravitating disks 
as a viscous process. We note that if $\rho\nu$ is constant in time
then increasing the density corresponds to reduction in the viscosity. 
This is perhaps not inconsistent with numerical simulations of disk
fragmentation which show the internal flow of high-density clumps
appear laminar \citep{gammie01}. However, once clumps form they may
effectively decouple from the background disk state, and thus no 
longer be described by the same prescription. 
%we are not concerned with non-linear evolution    
%{\bf is this true? KMK: maybe caveat helps?}

@@

\subsubsection{$\beta$ cooling}\label{beta_cool_model}
In our $\beta$ cooling model, the energy loss per unit volume is specified  
as an explicit function of the thermodynamic variables. 
A prototypical example is 
\begin{align}\label{beta_cool}
  \Lambda(\rho, T) =
  %  \frac{P}{(\gamma-1)T}\frac{\left(T-T_\mathrm{irr}\right)}{t_c}, 
  \frac{\mathcal{R}\rho}{(\gamma-1)}\frac{\left(T-T_\mathrm{irr}\right)}{t_c}. 
\end{align}
Recall  $\tcool = \beta\Omega^{-1}$ is 
the cooling timescale with $\beta$ a constant input parameter. Here, 
\begin{align}\label{tirr_def}
  \tirr = \theta T_\mathrm{eq}
\end{align}
is a reference temperature field parameterized as a constant 
fraction $\theta$ of the equilibrium temperature
$T_\mathrm{eq}$. Physically, $\tirr$ is the floor temperature set by,
for example, stellar or background irradiation. We will thus refer to
$\theta$ as the `irradiation level'. {\bf ASK} As we discuss in
\S\ref{2d_inviscid}, the limits $\theta \rightarrow 0, \theta
\rightarrow 1$ do not exactly correspond to the limits of
non-irradiated or irradiation-dominated disks as discussed in the
literature \citep{kratter11,rice11}. 
%stellar illum or envelope rad

For $\beta$-cooling given by Eq. \ref{beta_cool} we assume an 
equilibrium polytropic relation 
\begin{align} 
  P  =
%  \frac{c_{s0}^2\rho_0}{\Gamma}\left(\frac{\rho}{\rho_0}\right)^\Gamma,
  %  \equiv \frac{c_s^2(z)}{\Gamma}\rho,
c_{s0}^2\rho_0\left(\frac{\rho}{\rho_0}\right)^\Gamma,
\end{align}
where $\Gamma$ is the constant polytropic index and $\rho_0 =
\rho(z=0)$ is the equilibrium midplane density. 
The vertical structure is first obtained from 
Eq. \ref{vert_eq1}---\ref{vert_eq2}, %given strength of self-grav, via Q3d
then inserted into 
Eq. \ref{thermal_eq} to infer the required viscosity profile for
thermal equilibrium. If  $\mathcal{H}_\mathrm{ext}=0$, 
\begin{align}\label{alpha_beta_relation}
\alpha(z) = \frac{1}{(\gamma-1)\beta
   q^2}\frac{c_s^2(z)}{c_{s0}^2}\left(1 - \theta\right). 
%  \alpha(z) = \frac{1}{\Gamma(\gamma-1)\beta
%    q^2}\left(\frac{\rho}{\rho_0}\right)^{\Gamma-1} \left(1 - \theta\right). 
\end{align}

@@

Note that, although we assume standard beta cooling in formulating the
linear problem later, any explicit 
cooling function, say $\Lambda_1(\rho,T)$, can be mapped to
that obtained from Eq. \ref{beta_cool} by equating their linearized
forms, i.e. $\delta \Lambda \equiv \delta \Lambda_1$. This then 
defines the $\beta$ and $\theta$ parameters to be used in the 
framework we develop. We do this in \S\ref{2dppd} where we adopt a
more realistic beta cooling function for PPDs. In that case, $\theta$
may or may not represent a physical irradiation.

\subsubsection{Radiative cooling}\label{rad_cool}
A more realistic treatment of cooling considers energy transfer by radiative diffusion. Then
\begin{align}
  \Lambda &= \nabla\cdot\bm{F}_\mathrm{rad},\label{rad_cool1}\\
  \bm{F}_\mathrm{rad}   &= -\frac{16\sigma T^3}{3\kappa_d\rho}\nabla T, \label{rad_cool2}
\end{align}
where $\sigma$ is the Stefan-Boltzmann constant and 
$\kappa_d$ is the (dust) opacity. We adopt
\begin{align}\label{opacity_law}
  \kappa_d = \kappa_{d0}T^b,
\end{align}
and take the constant index $b=2$ as appropriate for the cold outer regions
of a PPD with ISM-like dust grains \citep{bell94}, but retain the general notation $b$
to keep track of the opacity.    
%We use radiative cooling when considering 3D PPDs in \S.

@@

While radiative cooling is arguably more realistic than
$\beta$-cooling, it generally implies a vertically
non-isothermal equilibrium disk, and increases the order of the
linearized equations. It formally applies to optically-thick 
disks, but it is possible to modify the flux function to account
for optically-thin disks \citep{levermore81}. However, this
complication is beyond the scope of this work.
